SFJ: An implementation of Semantic Featherweight Java

### 2020-04-22 conference/camera ready/background.tex
Add details of method types and structural/nominal discussion. Fix typos

A {program} $(\wt{L}, e)$ is a pair of a sequence of class declarations $\wt L$, giving rise to a class hierarchy as specified by the inheritance relation, and an expression $e$ to be evaluated. 
%
A class declaration $L$ specifies the name of the class, the name of the parent class it extends, its typed fields, the constructor $K$ and its method declarations $M$.
The constructor $K$ initializes the fields of the object by assigning values to the fields inherited by the super-class and to the fields declared in the current \textbf{this} class.
%
A method declaration $M$ specifies the signature of the method, namely the return type, the method name and the formal parameter as well as the body of the method.
Notice that in our theoretical development we use unary methods, without loss of generality: tuples of arguments can be modelled by an object that instantiates a ``special''class containing as fields all the needed arguments.
Expressions $e$ include variables, constants, field accesses, method invocations and object creations.

>>>>

A {program} $(\wt{L}, e)$ is a pair of a sequence of class declarations $\wt L$, giving rise to a class hierarchy as specified by the inheritance relation, and an expression $e$ to be evaluated.
%
A class declaration $L$ specifies the name of the class, the name of the parent class it extends, its typed fields, the constructor $K$ and its method declarations $M$.
The constructor $K$ initialises the fields of the object by assigning values to the fields inherited by the super-class and to the fields declared in the current \textbf{this} class.
%
A method declaration $M$ specifies the signature of the method, namely the return type, the method name and the formal parameter as well as the body of the method.
Notice that in our theoretical development we use unary methods, without loss of generality: tuples of arguments can be modelled by an object that instantiates a ``special'' class containing as fields all the needed arguments.
Expressions $e$ include variables, constants, field accesses, method invocations and object creations.

### 2020-04-22 conference/camera ready/paper.tex
Add details of method types and structural/nominal discussion. Fix typos

Orthogonally, for object-oriented languages there are two approaches to defining subtyping relations: the \emph{nominal} and the \emph{structural} approach \cite{Integrating,Useful}.
Nominal subtyping is based on {declarations} by the developer and is \emph{name}-based: ``$A$ is a subtype of $B$ if and only if it is declared to be so, that is if the class $A$ extends (or implements) the class (or interface) $B$''.
Structural subtyping instead is based on the \emph{structure} of a class, its fields and methods: ``a class $A$ is a subtype of a class $B$ if and only if the fields and methods of $A$ are a superset of the fields and methods of $B$, and their types in $A$ are subtypes of the types in $B$''.
Intutively, extending a class with more fields and methods results in a smaller set of inhabitants for that class. For example, the set of objects of a class \emph{Student} is smaller than the set of objects of a class \emph{Person} as each Student is a Person, but not the other way around. Hence, \emph{Student} is a structural subtype of \emph{Person}, even if it is not declared to be so.

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable. Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that the above types can be defined in Java, but as well are not inhabitable.

@@

Note that both approaches to subtyping have their drawbacks.
%However, it can be argued that structural subtyping is counterproductive in writing logically correct programs.
Consider two structurally equivalent class/record types $coordinate = [x:int, y:int, z:int]$ and $colour = [x:int, y:int, z:int]$.
While structurally they can be used interchangeably in our type system, their ``meaning'' is completely different.
On the other hand, with nominal subtyping in Java one can define an overridden method to perform the opposite logic to what the super class is expecting.
Therefore, both approaches leave an expectation on the developer to check what they are doing is correct, hence the integration of both in SFJ is the key to overcoming these drawbacks, as one can choose which approach to use for a given task.

Furthermore, we have extended FJ with method types, demonstrated below to show another possible implementation of the \emph{Polygons} \autoref{polygons}.
$$
    \synclass Diagonal \ \{
    \mathbf{int}\ diagonal(\textit{(getDiagonal: void$\rightarrow$int)}  \ shape)\{\ldots\}
    \}
$$

We instead only implement the \emph{getDiagonal} method on polygons that support it. We can then specify to only accept classes that implement the method \emph{getDiagonal}. If we pass an argument of type \emph{Triangle}, it is not in the computed set of subtypes, so we get a typechecking error at compile time.
We do not have to generate a relation for all methods in all classes as this is computationally unnecessary and instead
only generate the set of subtypes on encountering a method type.

>>>>

Orthogonally, for object-oriented languages there are two approaches to defining subtyping relations: the \emph{nominal} and the \emph{structural} approach \cite{Integrating,Useful}.
Nominal subtyping is based on {declarations} by the developer and is \emph{name}-based: ``$A$ is a subtype of $B$ if and only if it is declared to be so, that is if the class $A$ extends (or implements) the class (or interface) $B$''.
Structural subtyping instead is based on the \emph{structure} of a class, its fields and methods: ``a class $A$ is a subtype of a class $B$ if and only if the fields and methods of $A$ are a superset of the fields and methods of $B$, and their types in $A$ are subtypes of the types in $B$''.
Intuitively, extending a class with more fields and methods results in a smaller set of inhabitants for that class. For example, the set of objects of a class \emph{Student} is smaller than the set of objects of a class \emph{Person} as each Student is a Person, but not the other way around. Hence, \emph{Student} is a structural subtype of \emph{Person}, even if it is not declared to be so.

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable. Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that the above types can be defined in Java. These types can be inhabited by assigning \emph{null} to all fields, but are not useful in practise.

@@

Note that both approaches to subtyping have their drawbacks.
%However, it can be argued that structural subtyping is counterproductive in writing logically correct programs.
Consider two structurally equivalent class/record types $coordinate = [x:int, y:int, z:int]$ and $colour = [x:int, y:int, z:int]$.
While structurally they can be used interchangeably in our type system, their ``meaning'' is completely different.
Therefore, we lose some safety in our language as it is unlikely we would like to use a colour where we expect a coordinate, but in the eyes of the type-checker, it is correct.

Dardha \emph{et al.}\cite{Dardha2017} suggest a possible extension to the language by including a $nominal$ keyword that can be used in the class signature to indicate that this class should only be considered by the nominal component of the type system. This would allow us finer control over the type system and eliminate the previously mentioned problems.
However, being optional, we argue this would create more bugs than problems it would solve, due to easily forgetting the keyword.
For example with destructors in C++ classes,leaving out the virtual keyword of the parent class destructor in certain situations can lead to memory leaks.
Instead a \emph{structural} keyword could be introduced as an extension to the language which specifies that this class is allowed to have structural subtypes.
A class with this keyword is not added to the subtyping relation of other structurally similar classes which are not marked \emph{structural} as this would weaken the invariant conditions of that class.

On the other hand, with nominal subtyping in Java one can define an overridden method to perform the opposite logic to what the super class is expecting.
Therefore, both approaches leave an expectation on the developer to ensure what they are doing is correct, hence the integration of both in SFJ is the key to overcoming these drawbacks, as one can choose which approach to use for a given task.

Furthermore, we have extended FJ with \emph{method types}, demonstrated below to show another possible implementation of the \emph{Polygons} \autoref{polygons}.
$$
    \synclass Diagonal \ \{
    \mathbf{int}\ diagonal(\textit{(getDiagonal: void$\rightarrow$int)}  \ shape)\{\ldots\}
    \}
$$

We define the method type as accepting any type and its subtypes which implement a method called \emph{diagonal} which is a method from the type \emph{Void} to a \emph{Double}.

In order to type-check the value passed to such an argument, we can at compile-time build a collection of types $\{type_{1},\ type_{2},\ \ldots\}$ which have this method.
We can do this the simply iterating over the same list of classes in the program as when we were creating our subtyping relation and checking for the presence of the required method.
The set of accepted types would then be the union of all their corresponding subtypes $([\![type_{1}]\!]_{B} \cup [\![type_{2}]\!]_{B} \cup\ \ldots)$.

However, calculating this collection of types for each method of every class would be computationally unnecessary as only few of these would be used for method types.
Therefore we only compute them on demand during type-checking when we come across such a type.

We can therefore use method types to statically include or exclude a portion of our type hierarchy.
However unlike the use of interfaces such as in the suggested Java solution to the Polygons problem \autoref{polygons}, the values that can be accepted by a method type do not have to be related to each other in any way in the class hierarchy.
This becomes increasingly useful if we have a legacy system as we can still accept all the classes that have defined \emph{Diagonal} methods without having to go back and add interface implementations.
